---
title: "Data 606 Final Project"
author: "Cameron Smith"
date: "12/3/2020"
output: 
  ioslides_presentation:
    widescreen: true
    smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r include=FALSE}
library(tidyverse)
library(ggplot2)
library(cowplot)
library(kableExtra)
library(countrycode)
# Load data from Github
file = "https://raw.githubusercontent.com/cwestsmith/cuny-msds/master/datasets/wvsdata_606project.csv"
df <- read.csv(file, header=TRUE)

# Create data frame using with more readable column names and excluding the row number
names(df) <- c("row_num", "country_code", "level_happiness", "life_satisfaction", "num_children")
df <- df %>% select("country_code", "num_children", "life_satisfaction")

# Remove rows with NA values
df_new <- na.omit(df)

# Add a variable that matches the region to each country name
df_new <- df_new %>% 
  mutate(region = countrycode(sourcevar = df_new$country_code, origin = "iso3c", destination = "region"))

# Check for NA values in the new column, which indicates 0 issues
df_new %>% filter(is.na(region) == TRUE)
```

## Overview

As someone about to enter the world of parenthood I thought it would be interesting to look at the potential relationship between having children and one's level of happiness.  The specific research question I had *planned on* asking is: 

***Is there a difference in levels of happiness between non-parents and parents?***

However, mid-way through my original analysis I realized that there were issues with the data which rendered it less useful so I re-framed the question as:

***Is there a difference in levels of life satisfaction between non-parents and parents?***

**Null hypothesis:** There is no difference in average levels of life satisfaction between non-parents and parents

**Alternative hypothesis:** There is some difference in average levels of life satisfaction between non-parents and parents

## Data / Data Collection

The data is "Wave 7" from the World Values Survey (WSV), and the details - including a link to where the full data set can be downloaded, is below.  However, please note that the date used in this file is an extract I made from that original data set that only includes data for the variables relevant for this project.  The original file includes 536 separate variables.

The dataset currently (the website notes that it is still a work in progress) spans 77 countries and includes 69,578 observations covering the years 2017 to 2020.

**Citation:**

WVS wave 7 (2017-2020): Haerpfer, C., Inglehart, R., Moreno, A., Welzel, C., Kizilova, K., Diez-Medrano J., M. Lagos, P. Norris, E. Ponarin & B. Puranen et al. (eds.). 2020. World Values Survey: Round Seven (2017 to 2020) Country-Pooled Datafile. Madrid, Spain & Vienna, Austria: JD Systems Institute & WVSA Secretariat [Version: http://www.worldvaluessurvey.org/WVSDocumentationWV7.jsp].

## Key Variables 

* The dependent (response) variable is level of life satisfaction, which is technical a categorical (ordinal) variable with values ranging from 1 to 10 but per discussion with the professor will be treated as a quantitative variable for this project.\
* The main independent (explanatory) variable will be whether the respondent is a parent, which is a binary variable (is parent, or is not parent).\
* The secondary independent variable will be the country they live in, which is categorical (nominal).\

```{r}
glimpse(df_new)
```

## Exploring the Data

```{r message=FALSE, warning=FALSE}
# How many levels of life satisfaction are in the dataset?
df_new %>% group_by(life_satisfaction) %>% summarize("num_rows" = n()) %>% kable()

```

## Exploring the Data

```{r message=FALSE, warning=FALSE}
# Histogram of number of children
hist(df_new$num_children, xlab = "Number of Children", main = "Histogram: Number of Children")
```

## Exploring the Data

```{r message=FALSE, warning=FALSE, include=FALSE}
# Mean of life satisfaction level for top 10 countries w/ mean number of children
df_summarized <- df_new %>% group_by(country_code) %>% 
  summarize("avg_life_satisfaction" = mean(life_satisfaction), "avg_num_children" = mean(num_children))
```

Summary by avg life satisfaction, descending
```{r echo=TRUE, message=FALSE, warning=FALSE}
top_n(arrange(df_summarized, desc(avg_life_satisfaction)), 10)
```

## Exploring the Data

Summary by avg num of children, descending
```{r echo=TRUE, message=FALSE, warning=FALSE}
top_n(arrange(df_summarized, desc(avg_num_children)), 10)
```


```{r message=FALSE, warning=FALSE, include=FALSE}
# Separate data for parents and non-parents
parents_df <- df_new %>% filter(num_children > 0)
nonparents_df <- df_new %>% filter(num_children == 0)
```

## Exploring the Data

Box Plots of little use based on ordinal variable, though differing IQRs are interesting

```{r echo=TRUE, message=FALSE, warning=FALSE}
# Boxplots of life satisfaction levels - parents vs non-parents
boxplot(parents_df$life_satisfaction, nonparents_df$life_satisfaction, names = c("parents", "non-parents"), ylab = "Life Satisfaction", main = "Life Satisfaction (Parents vs Non-Parents)")
```


```{r message=FALSE, warning=FALSE, include=FALSE}
parents_df %>% group_by(life_satisfaction) %>% summarise(num_observations = n())
nonparents_df %>% group_by(life_satisfaction) %>% summarise(num_observations = n())
```

## Summary Statistics and Inference

To determine whether there is a significant difference between the two population means (parents and non-parents), a T-test was used.

As a first step, we need to evaluate the two Central Limit Conditions:

* **Independence, extended:** The WVS website says 'All countries employed random probability representative samples of the adult population', so this condition is satisfied.

* **Normality:** The sample size is significantly larger than 30 for each group, as detailed below, so assuming there are no clear outliers this condition is satisfied.

## Summary Statistics and Inference

To further check normality histograms and QQ plots were created for each group:

```{r echo=FALSE}
# Histograms to check normality condition for both groups
hist_nonparents <- ggplot(data=nonparents_df, aes(life_satisfaction)) + geom_histogram(binwidth = 1)
hist_parents <- ggplot(data=parents_df, aes(life_satisfaction)) + geom_histogram(binwidth = 1)

# QQ plots to test normality of data for both groups
qq_nonparents <- ggplot(nonparents_df, aes(sample = life_satisfaction)) +
  stat_qq() +
  stat_qq_line()

# QQ plot to test normality of data for parents
qq_parents <- ggplot(parents_df, aes(sample = life_satisfaction)) +
  stat_qq() +
  stat_qq_line()

plot_grid(hist_nonparents,
          hist_parents,
          qq_nonparents,
          qq_parents,
          labels = c('Non-Parents','Parents'),
          label_x = 0.2,
          ncol = 2)
```

## Summary Statistics and Inference {.flexbox .vcenter}

After confirming the conditions for inference have been met, summary statistics and a t-test can be performed.  The t-test was first done manually and then done again via an r function to compare / confirm results.

```{r message=FALSE, warning=FALSE, include=FALSE}
# Find the means of each group
nonparents_mean <- mean(nonparents_df$life_satisfaction)
parents_mean <- mean(parents_df$life_satisfaction)
difference_mean <- parents_mean - nonparents_mean

# Find the standard deviation of each group
nonparents_sd <- sd(parents_df$life_satisfaction)
parents_sd <- sd(nonparents_df$life_satisfaction)

# Find 'N' (number of observations) for each group
nonparents_n <- nonparents_df %>% summarize(n()) %>% as.numeric()
parents_n <- parents_df %>% summarize(n()) %>% as.numeric()

# Degrees of freedom, take the smaller of the two observation counts and subtract 1
degfreedom <- if_else(nonparents_n < parents_n, nonparents_n - 1, parents_n - 1)

# Calculate the standard error
standard_error <- sqrt(
  (nonparents_sd ^ 2 / nonparents_n) +
    (parents_sd ^ 2 / parents_n)
)

# Calculate the t-statistic
tstat_manual <- difference_mean / standard_error

# Find the 95% confidence interval for the difference in means
interval_low <- difference_mean - standard_error
interval_high <- difference_mean + standard_error

# The p-value is smaller than 0.05, so we reject the null hypothesis
pvalue_manual <- 2*pt(-abs(tstat_manual), df=degfreedom)
result <- if_else(pvalue_manual > 0.05, "fail to reject", "reject")
```

```{r}
# The means seem to be fairly similar, although non-parents are slightly higher
# ...But is this difference statistically significant?
cat("Nonparents:", "\n Mean:",round(nonparents_mean,2), "\n Number of Observations (N):", nonparents_n, "\n Standard Deviation:", nonparents_sd, "\n\n")
cat("Parents:", "\n Mean:",round(parents_mean,2), "\n Number of Observations (N):", parents_n, "\n Standard Deviation:", parents_sd, "\n\n")
cat("Standard Error:", standard_error, "\nDegrees of Freedom:", degfreedom)
```

Results of manually performed t-test
```{r}
cat("P-value:", pvalue_manual, "\nT-statistic:", tstat_manual, "\n95% Confidence Interval is between", round(interval_low, 3), "and", round(interval_high, 3), "\nThe p-value is", round(pvalue_manual,3), "so we", result, "the null hypothesis")
```

## Summary Statistics and Inference

Same t-test but using built-in t.test function, which confirms the same result

```{r echo=TRUE}
autotest <- t.test(parents_df$life_satisfaction, nonparents_df$life_satisfaction)
autotest
```

## Taking Another Look

```{r include=FALSE}
# Average life satisfaction by # of children
avg_satisfaction_parents <-
  parents_df %>% 
  group_by(num_children) %>% 
  summarize("num_rows" = n(), "avg_life_satisfaction" = mean(life_satisfaction))

# Average levels of life satisfaction by region, descending
avg_satisfaction_parents_region <- parents_df %>% 
  group_by(region) %>% 
  summarise(avg_life_satisfaction = mean(life_satisfaction))
```

I performed some more analysis to see if perhaps there was a difference in regions, which there seems to be.

The data seems to indicate that satisfaction is quite a bit better after 13 children, and Latin America seems to be the highest scoring region.
```{r echo=FALSE, message=FALSE, warning=FALSE}
plot_avg <- ggplot(data=avg_satisfaction_parents, aes(x=num_children, y=avg_life_satisfaction)) +
    geom_bar(stat="identity") +
    xlab("Number of Children") + ylab("Avg. Life Satisfaction")
plot_region <- ggplot(data=avg_satisfaction_parents_region, aes(x=region, y=avg_life_satisfaction)) +
    geom_bar(stat="identity") + coord_flip() +
    xlab("Region") + ylab("Avg. Life Satisfaction")
plot_grid(plot_avg,
          plot_region,
          label_x = 0.2,
          ncol = 2)
```


## Another T-Test: Is This Normal?
Another normality check, this time for parents w/ 14+ children

```{r include=FALSE}
parents_happy <- parents_df %>% filter(num_children >= 14)

avg_satisfaction_happyparents <-
  parents_happy %>% 
  group_by(num_children) %>% 
  summarize("num_rows" = n(), "avg_life_satisfaction" = mean(life_satisfaction))

# Check normality assumption again, though technically satisfied with 31 observations.  Not very normal.
plothappyhist <- ggplot(data=parents_happy, aes(life_satisfaction)) + 
  geom_histogram(binwidth = 1) + 
  ggtitle("Life Satisfaction: 14 or more children") +
  xlab("Life Satisfaction") + ylab("Number of Observations")

plotchildrencount <- ggplot(data=avg_satisfaction_happyparents, aes(x=num_children, y=num_rows)) +
    geom_bar(stat="identity") +
    ggtitle("# Observations") +
    xlab("Number of Children") + ylab("Number of Observations")
```


```{r echo=FALSE}
plot_grid(plothappyhist,
          plotchildrencount,
          label_x = 0.2,
          ncol = 2)
```

## Another T-Test

Running another t-test, with the filtered group, yielded different results.  P is < 0.5 and thus the null hypothesis cannot be rejected.

```{r echo=TRUE}
# Run another t-test, this time focused using data for parents with 14 or more children
t.test(parents_happy$life_satisfaction, nonparents_df$life_satisfaction)
```

## One Last Try
What if we changed the approach a bit and use a one tailed t-test?  

Surprise!  Based on this test, with a p value of under .05 we can reject the null hypothesis and say that parents with 14 or more children are on average more satisfied with their lives than non-parents.

```{r}
t.test(parents_happy$life_satisfaction, nonparents_df$life_satisfaction, alternative = "greater")
```

## Conclusion

* There is a statistically significant difference in levels of life satisfaction between parents and non-parents
* The difference appears to be positive
* Ordinal variables are not ideal for quantitative analysis
* One-tailed tests can be deceiving
* ANOVA would perhaps have been a better test



