---
title: "Data 607 - Week 9 Homework"
author: "Cameron Smith"
date: "`r Sys.Date()`"
output: 
  html_document:
  toc: true
---

```{r load-packages, message=FALSE}
library(tidyverse)
library(knitr)
library(httr)
library(jsonlite)
```

# Introduction / description

For this assignment our task was to choose one of the New York Times APIs, construct an interface in R to read the JSON data, and transform it into an R DataFrame.

## Overview of approach

I chose the "Most Popular" API which makes available information on the most popular NY Times articles.  I retrieved info for the last day, though it is also possible to increase the number by modifying the API call.  This required first setting up a NY Times developer account, and from there selecting the API I want to use for which an API key was assigned.

Link to more information:
https://developer.nytimes.com/docs/most-popular-product/1/overview

## Retrieve the data

The data is retrieved using the GET function from the httr package, which retrieves JSON data.  It is then converted via the fromJSON function.

```{r echo=TRUE}
# Send GET request
res = GET("https://api.nytimes.com/svc/mostpopular/v2/viewed/1.json?api-key=9OzGBXzGGQI4Ql8IEMXVM2CiMTPp63T2")

# Confirm it was successful - code 200 is desired
cat("The status returned is:", res$status_code)

# Convert JSON data
data <- fromJSON(rawToChar(res$content))

#Check to see what came back
names(data)

```

## Structure the data

The key data can be extracted and placed in a data frame by selecting it as follows.

```{r}
# Create the data frame
df <- data.frame(data$results, stringsAsFactors = FALSE)
```

## View the data

```{r}
# View summary of top articles
df %>% 
  select(title, section, published_date) %>% 
  rename("Title" = title, "Section" = section, "Published Date" = published_date) %>% 
  kable

# Confirm that the data is in a data frame (objective of the assignment)
truefalse <- is.data.frame(df)
cat("Is this a dataframe?", truefalse)

```

# Conclusion

Using a combination of the httr and jsonlite packages we can retrieve, format and use data from APIs - in this case the NY Times.