---
title: "Data 607 - Project 1"
author: "Cameron Smith"
date: "9/17/2020"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This project was focused on the analysis of chess tournament data.  Using cross-tables from a text file we needed to extract the key data using regex, transform it into a usable form, and then analyze specific elements including in particular the pre-ratings of each player and their opponents.

### Load library and read File
The Tidyverse is used in order to leverage its filtering and regex capabilities.  For reproducibility I placed the text file in Github, read directly into R.

```{r}
# Load required libraries
library(tidyverse)

# Get the data
raw_data <- readLines("https://raw.githubusercontent.com/cwestsmith/cuny-msds/master/datasets/tournamentinfo.txt")
```

### Extract key fields

Key fields were extracted using the str_extract_all function.  The extracted data was then put into a data frame.

```{r}
# Extract the required data using regular expressions
player_num <- as.numeric(unlist(str_extract_all(raw_data,"(?<=\\s{3,4})\\d{1,2}(?=\\s)")))
player_name <- unlist(str_extract_all(raw_data,"(?<=\\d\\s\\|\\s)([A-z, -]*\\s){1,}[[:alpha:]]*(?=\\s*\\|)"))
player_state <- unlist(str_extract_all(raw_data, "[[:upper:]]{2}(?=\\s\\|)"))
total_pts <- as.numeric(unlist(str_extract_all(raw_data, "(?<=\\|)\\d\\.\\d")))
player_pre_rat <- as.numeric(unlist(str_extract_all(raw_data, "(?<=R:\\s{1,2})(\\d{3,4}(?=\\s))|(\\d{3,4}(?=P\\d{1,2}\\s*-))")))

# Take the extracted data and put it into a data frame
processed_data <- data.frame(player_num, player_name, player_state, total_pts, player_pre_rat)

# Check the data frame's structure to make sure it is as intended (i.e. number columns are numeric, character columns are character, etc..., and that it has the correct number of rows)
str(processed_data)
```

### Create list of opponent player numbers

The next step was to extract the opponent player numbers, including opponents for which matches did not materialize (bys and forfeits, for example) which brought with it some challenges due to the format of the file.  A combination of approaches was used to extract the data.

```{r}

# Had some initial challenges doing this with only regex so to make it simpler and a bit more robust I created a new list that only had included the relevant rows from the raw data file.
secondary_rows <- raw_data[seq(5, 196, 3)]
opponent_num <- as.numeric(unlist(str_extract_all(secondary_rows, "(?<=\\|(W|L|D)\\s{2,3})[[:digit:]]{1,2}(?=\\|)|((?<!->)(?<=\\|(U|H|B|X))\\s{4}(?=\\|))")))
```

### Calculate the average pre chess rating (PCR) of each player's opponents

For this step I used a for loop to take sequences of each 7 entries (corresponding with the 7 chess matches for each person), add them together for the total opponents' score, then divide them by the number of entries that were not NA in that sequence to find the average.

```{r}

# Create matrix to store data calculated in the for loop.  Pre-populating values with NA for more efficient processing in R.
pcr_matrix <- matrix(data = NA, nrow = 64, ncol = 2)

# Assign readable names for the matrix
colnames(pcr_matrix) <- c("total_opp_pcr", "avg_opp_pcr")

# Initialize a variable to be used as a counter in the for loop to fill the corresponding matrix row
row_counter <- 0

# Start of for loop
for(i in seq(from=1, to=length(opponent_num)-6, by=7)){
  row_counter <- row_counter + 1
  
# Perform a lookup of each competitor's score based on their player number and add the up for each row (corresponding to each sequence of 7 data points, w/ value from for loop serving as row 'anchor')
  pcr_matrix[row_counter, 1] <- (sum(subset(processed_data$player_pre_rat, processed_data$player_num %in% opponent_num[seq(from=i, to=i+6, by=1)])))
  
# Calculate the average score for each row, excluding missing entries
  pcr_matrix[row_counter, 2] <- pcr_matrix[row_counter, 1] / length(subset(opponent_num[seq(from=i, to=i+6, by=1)],!is.na(opponent_num[seq(from=i, to=i+6, by=1)])))
  
}
# End of for loop

# Verify that matrix was processed properly by looking at the first few rows of output
head(pcr_matrix, 5)
```

## Final tidying up

The average PCRs were then rounded up to the nearest full point and added to the data frame with the other processed data.

```{r}

# Round the figures to the nearest whole number
pcr_matrix[, 2] <- round(pcr_matrix[,2], digits = 0)

# Add average scores to data frame with other processed data and rename for readability
processed_data <- cbind(processed_data, pcr_matrix[, 2])
processed_data <- rename(processed_data, avg_opp_pcr = `pcr_matrix[, 2]`)
```

## Export of .csv file

The final data frame of processed data was then exported into a .csv file for ease of reference.  A platform independent path has been used to ensure full reproducibility.
```{r}

# Get working directory path
path <- getwd()

# Export file to working directory.  The file.path function has been used to ensure platform independence (i.e. take into account the different path syntaxes for various operating systems)
write.csv(processed_data, file.path(path, "chess_processed_data.csv"))

```

# Conclusion

Using regex in combination with various Tidyverse functions it was possible to extract the data needed from a text file with a somewhat irregular structure.  That data could then be fed into a data frame for ease of analysis, using a for loop to process complex elements and minimize the lines of code required.  The complete dataframe with all processed data could then be exported into a .CSV file for further reference if/as needed.  A sample of the complete dataframe is below for reference.

```{r}
head(processed_data, 5)
```

